Modeling the movement dynamics of hippocampal replay reveals slower spatially coherent trajectories

### 2020-10-01 Article.tex
Update on Overleaf.

\title{Revealing the movement dynamics of hippocampal replay using state space modeling}
\shorttitle{Replay Dynamics}

@@

%TC:break Abstract
%the command above serves to have a word count for the abstract
\begin{abstract}
Representations of past and possible future experiences play a critical role in memory and decision-making processes. The hippocampus is known to express such representations during sharp-wave ripple (SWR) events, when ensembles of hippocampal place cells can be activated in time-compressed sequences consistent with prior experiences. Efforts to understand these "replay" events have typically imposed stringent criteria on SWR event significance, and thus analyze only a small subset of events (\texttildelow10-45\%). As a result, we lack a clear understanding of both the prevalence and the range of representational dynamics associated with replay. Here, we develop a state space model that uses a combination of movement dynamics to capture the spatial content and time evolution of replay during SWR events. Application of this model reveals that the large majority of SWR events contain coherent spatial content that evolves in accordance with a surprisingly wide range of mental movement speeds.

@@

We find that the majority of SWRs contain spatially coherent content; that is, the replay trajectories have no large jumps. Surprisingly, while the expected high speed replay events were identified, the most common category of events expressed representations that moved at much slower speeds, more consistent with real world animal movement, suggesting that high speed replay is not the most common manifestation of generative activity. These findings illustrate the power of state space models and provide a new understanding of the nature of hippocampal replay.

\section*{Results}
\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure1/Figure1_v6}
\caption{The model can capture different sequence dynamics on simulated data. \textbf{(A)} We construct a firing sequence of 19 simulated place cells that exhibits three different movement dynamics. For the first 60 ms, one cell fires repeatedly, representing one stationary location. For next 190 ms, the cells fire in sequence, representing a rapid continuous trajectory along the virtual track. Finally, for the last 30 ms, cells fire randomly, out of spatial order, representing an fragmented spatial sequence. \textbf{(B)} Like the standard decoder, the state space model uses estimates of cells' place fields from when the animal is moving and combines them with the observed spikes in (A) to compute the likelihood of position for each time step. \textbf{(C)} The prediction from the neural data is then combined with an explicit model of each movement dynamic, which determines how latent position can change based on the position in the previous time step. We show the probability of the next position bin for each movement dynamic model (color scale). Zero here represents the previous position. \textbf{(D)} The probability of remaining in a particular movement dynamic versus switching to another dynamic is modeled as having a high probability of remaining in a particular dynamic with a small probability of switching to one of the other dynamics at each time step. \textbf{(E)} The state space model uses the components in A-D over all time to decode the joint probability of latent position and dynamic state. The joint probability can be summarized by marginalizing over latent position (left panel) to get the probability of each dynamic state over time. The summed posterior across states also provides an estimate of latent position over time (right panel). \textbf{(F)} The probability of each dynamic depends heavily on the speed of the trajectory, as we show using a range of simulated spiking sequences each moving at a constant speed. Each dot corresponds to the average probability of that dynamic for a given constant speed sequence. We use a 0.8 threshold (dotted line) to classify each sequence based on the dynamic or dynamics which contribute maximally to the posterior (shaded colors). 
}
\label{1}
\end{figure*}
\subsection*{Overview of the model}
We begin with an application of the model to simulated data, both to validate the model and to provide intuition (Figure \ref{1}). We simulated 19 Poisson spiking cells with Gaussian place fields on a 180 cm virtual linear track. Each place field has a 36 cm variance and a 15 Hz peak firing rate, which is spaced every 10 cm along the virtual track. We then construct an 180 ms spiking sequence and apply our decoding algorithm in Figure \ref{1}A. For the first 60 ms of this sequence, a single place cell fires repeatedly, resulting in the extended representation of a single location. For the second 60 ms of the sequence, the cells fire in sequential spatial order, representing a fast moving trajectory across the virtual linear track. For the last 60 ms of the sequence, the cells fire in an incoherent spatial order. These three firing patterns represent three different types of sequence dynamics that could be expressed during replay events, which we call stationary, continuous, and fragmented, respectively. The goal of our model is to characterize SWRs in terms of a mixture of these three dynamics at every time point.

@@

Next, we specify movement dynamic models for how latent position---the "mental" position of the animal represented by the cells---evolves over time for a given dynamic (Figure \ref{1}C). We do this by defining a state transition matrix, which defines how the latent position can move from the previous time step. Previous findings suggest that replay may exhibit at least three distinct types of movement dynamics: stationary \cite{JaiDistincthippocampalcorticalmemory2017, FarooqEmergencepreconfiguredplastic2019}, rapid continuous \cite{DavidsonHippocampalReplayExtended2009}, and fragmented, which could correspond to both extended spatially incoherent representations and representations that jump from one position to another in a short time period \cite{PfeifferAutoassociativedynamicsgeneration2015}. We therefore define state transition models to capture each of these movement dynamics.

@@

Once we have specified the data and the movement dynamics models, we have fully specified the state space model. We use acausal decoding, meaning that we use all information from the past and future spikes, to estimate the joint posterior probability of position and dynamic (see Methods). With this, we can summarize the resulting posterior with two quantities: the probability of each movement dynamic over time (by integrating out position) and the probability of latent position over time, irrespective of movement dynamic (by summing over the movement dynamics; Figure \ref{1}F, left and right plot respectively).

An examination of the two summaries shown in Figure \ref{1}F reveals that that the model successfully captures the dynamics of the population spiking activity in Figure \ref{1}A. The stable firing of the one active neuron indicates a stationary representation, and accordingly, the probability of the stationary movement dynamic is high at the beginning of the simulated replay. A change in the data then drives a rapid transition to the continuous movement dynamic, reflecting the trajectory-like spatially sequential spiking from the population of cells. Subsequently, as the population activity becomes spatially incoherent, the fragmented movement dynamic dominates for the last third of the simulated event.

The specification of these movement dynamics allows the model to capture a wide range of movement speeds for the latent position. To demonstrate this, we applied the model to 10,000 simulated trajectories, each trajectory proceeding at a constant speed (Figure \ref{1}G) from 1 cm/s to 10,000 m/s. From this, we can see that not only are there regions of speed that correspond to our three movement dynamics being highly probable (where we define highly probable to be greater than or equal to 0.8 probability), there are intermediate speeds where two of the dynamics exhibit relatively high probability; and where the sum of two of the dynamics' probabilities exceeds 0.8. In this manuscript, we will refer to these as mixture dynamics. For example, when the stationary state has a probability of 0.6 and the continuous has a probability of 0.4, we call this a stationary-continuous mixture (light blue, Figure \ref{1}G). By using this classification scheme, we can characterize the speed or set of speeds in a replay by its movement dynamics. To summarize, there are three movement dynamic models, each of which is assigned a probability over time. We classify these probabilities as being part of one of five speed categories: Stationary, Stationary-Continuous-Mixtures, Continuous, Fragmented-Continuous Mixtures, and Fragmented.

@@

As mentioned above, the model is robust to the choice of probability of staying in a state. To demonstrate this we decoded the spiking sequence in Figure \ref{1}A with different probabilities of staying in the same state versus switching to another state (Figure \ref{fig:Figure1-Figure supplement 1}). We found that for a large range of plausible probabilities of staying in one of the dynamics over a single 2 ms time step (0.9 - 0.9999), the model still correctly identified the dynamic with high probability.

@@

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure3/Figure3_v5}
\caption{A-F. Six examples of SWRs that would not be well-characterized by using the standard decoder. Figure conventions are the same as in Figure \ref{2}. Filtered SWR ripple (150-250 Hz) trace from the tetrode with the maximum amplitude displayed above each example. \textbf{(A)} A SWR where the decoded position starts moving down the center arm away from the animal's position at the center well, slows down, and returns back. \textbf{(B)} A SWR where the decoded position persistently stays at the choice point (open circle) while the animal remains at the left well. \textbf{(C)} A SWR where the decoded position begins with stationary representation of the left well, then jumps to the middle of the right arm and proceeds up the right arm to the right well. \textbf{(D)} A SWR where the decoded position begins with stationary representation of the left well, jumps to the center arm, proceeds away from the center well, jumps to the right arm, proceeds back toward the center well, and then becomes fragmented. \textbf{(E)} A SWR where the decoded position begins in the left arm and persists at the choice point. \textbf{(F)} A SWR where the decoded position starts in the left arm toward the choice point, jumps to the right arm and proceeds back toward the choice point. \textbf{(G)} Each dot represents the percentage of SWRs for each day for that animal. A SWR is included in the numerator of the percentage if at any time it includes the classifications listed below the column. The denominator is listed in the x-axis label.
}
\label{3}
\end{figure*}

We next sought to validate the model on real hippocampal data. We fit the place fields of 31 cells recorded in hippocampus while the rat was performing a spatial alternation task on a W-shaped track, and then applied the decoding algorithm to a SWR with sequential population activity (Figure \ref{2}A, top panel). As expected, we observed that the probability of being in the continuous dynamic is high throughout this event, but the probability of being in a stationary state was also noticeable at the beginning and end of the SWR (Figure \ref{2}A, middle panel). Using our speed classification scheme, this means that the speed of the replay starts slower---as a mixture of continuous and stationary dynamics---and then speeds up and slows down again. This is also evident in the posterior probability of linear position over time. This shows that the replay most likely travels down the center arm and up the right arm (Figure \ref{2}A, bottom panel). We can also see this when we project the maximum of the posterior of this trajectory (the most probable "mental" position) to 2D (Figure \ref{2}C) to better see the spatial trajectory on the maze. Importantly, when we apply the same model using the 2D position of the animal, we get a similar result as when we use the 1D linearized position (Figure \ref{fig:Figure2-Figure supplement 1}A).

@@

As expected, and providing further support for the validity of the method, we observed many replays that were classified as continuous throughout the entirety of the SWR (Figure \ref{fig:Figure2-Figure supplement 2}A-C). However, we also observed many events with spatially coherent content that did not have this structure. For example, we observed trajectories that started in one direction and reversed back to the original position (Figure \ref{3}A, Figure \ref{fig:Figure3-Figure supplement 1}), trajectories that remained fixed in one position (Figure \ref{3}B), trajectories that jumped between arms and between dynamics (Figure \ref{3}C-D, Figure \ref{fig:Figure3-Figure supplement 1}A, E, F), and trajectories that were spatially incoherent throughout the SWR (Figure \ref{fig:Figure3-Figure supplement 1}D).

Using a 0.8 threshold, we were able to classify 89\% (23382 of 26160) of SWRs as containing at least one of the three dynamics or common state mixtures. To ensure that this depended on the sequential firing of spatially tuned cells in the hippocampus, we resampled position with replacement for two recording sessions, shuffling the relationship between spiking and spatial information and then fitting the encoding model. We then decoded the same SWR events containing the same spikes. Only 9\% of the SWRs were classified in the shuffled datasets, a value that was significantly less than that seen for the real data (p=0.02 for recording session 1, p=0.02 for recording session 2, Figure \ref{fig:Figure3-Figure supplement 2}).

@@

To better compare our findings to previous work, we quantified the percentage of classified SWRs that contained continuous content, as would typically be analyzed when using the standard decoder. Here our findings were consistent with previous reports: in our datasets, 4511 of 23382 or 19\% of classified SWRs had some continuous content (Figure \ref{3}G, 17\% of all SWRs). Thus, focusing on only high speed trajectories misses a large fraction of events where there is evidence for spatially coherent, but not continuous content.

@@

Our model is specified in the context of a latent position associated with different movement dynamics. This specification allows us to not only to classify events in terms of their dynamics, but also to quantify the model's certainty in each position estimate at each moment in time given the model parameters. To do so we can compute the cumulative spatial extent of the 95\% highest posterior density (HPD) region of the posterior linear position. Larger values of the cumulative HPD extent indicate the model is less certain about position, because the HPD values are distributed over more of the track at a given time point. In contrast, lower values indicate that the model is more certain about the estimate of position because the extent of the HPD is more concentrated and covers less of the track. Thus, the HPD provides a complementary measure of spatial coherence, and evaluating it allows us to verify that the events we defined as spatially coherent correspond to events where there is high certainty around the position estimates. Although this is certainly influenced by the dynamics imposed on the model via the movement state transition matrix, because the model is driven by the data, it does not necessarily have to be the case that the

@@

\subsection*{Many replay trajectories have slower movement speeds}
\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure5/Figure5_v4}
\caption{
Quantification of the SWRs by dynamic. \textbf{(A)} UpSet plot \cite{LexUpSetVisualizationIntersecting2014}---which is similar to a Venn diagram with more than 3 sets---of the most common sets of classifications within each SWR. Each row represents a classification and each column represents a set of classifications, where filled-in black dots with an edge between the dots indicates that multiple classifications are present in the SWR (at temporally distinct times). The sets are ordered by how often they occur as indicated by the bar plot above each category. The total percentage of each classification is indicated by the rotated bar plot on the left. \textbf{(B)} Box plot of the duration of each dynamic category within a SWR. The box shows the interquartile range (25-75\%) of the data and the whiskers are 1.5 times the interquartile range. Points outside of this are labeled as diamonds. \textbf{(C)} Box plot of the average distance between the latent position to the animal's position for each classification within the SWR. \textbf{(D)} Box plot of the average speed of the classification within the SWR (excluding classifications < 20 ms).\textbf{(E)} Kernel density estimate of the position of stationary trajectories on the W-track at least 30 cm away from the animal's position. The shaded region represents the density estimate while the black hatch marks represent the observed non-local stationary positions. \textbf{(F)} Box plot of the average tetrode multiunit spike rates for each dynamic category within each SWR (excluding classifications < 20 ms).
}
\label{5}
\end{figure*}

@@

Previous work reported that less than half of SWRs or high-population-activity events contained replay \cite{KarlssonAwakereplayremote2009, FosterReversereplaybehavioural2006, DavidsonHippocampalReplayExtended2009}. This could lead one to the assumption that most events are not spatially coherent or meaningful. Our results indicate that this is not the case: in our dataset, the large majority of SWRs (85\%) contained spatially coherent content, as defined by some combination of stationary or continuous dynamics. These events included decoded representations that were highly spatially concentrated, confirming their spatial coherence. This was in stark contrast to unclassified events or events that included fragmented dynamics, which were associated with highly distributed decoded spatial representations.

@@

Second, we used clusterless decoding, which enabled us to take advantage of information from neurons farther away from the tetrodes and not just those that could be cleanly and somewhat subjectively separated \cite{ChenBayesiannonparametricmethods2016, KloostermanBayesiandecodingusing2014, DengRapidclassificationhippocampal2016}. Third, we built an explicit, but flexible model that allowed us to identify not just one dynamic, but multiple dynamics consistent with different speeds of motion. And we use these dynamics to describe the SWR events, rather than declaring the events as significant or not. Fourth, our model avoids assumptions about temporal bins and does not make assumptions about the linear progression of events. Instead, the model can use any size of bin and allows for movement in any direction. Finally, we used an acausal estimator of latent position, which allowed us to identify the specific timing of transitions in dynamics at a fine timescale. Our method allows us to identify periods of slower moving events that

@@

\subsection*{Mixture probability of the dynamics}
The mixture probability of the dynamics is a quantity that indicates how much that dynamic is contributing to the posterior at a given time. We estimate the mixture probability of the dynamics by integrating out position from the joint posterior:
$$Pr(I_{k} \mid O_{1:T}) = \int p(x_{k}, I_{k} \mid O_{1:T}) dx_{k}$$
As in other calculations, the integral is approximated with a Riemann sum.

@@

\subsection*{Classification of Dynamics}
We used a threshold of 0.8 to classify the probability of each state $Pr(I_{k} \mid O_{1:T})$ into 5 categories. Time periods during sharp wave ripples are labeled as Stationary, Continuous, or Fragmented when the probability of each individual state is above 0.8. Time periods are labeled as Stationary-Continuous-Mix or Fragmented-Continuous-Mix when the sum of Stationary and Continuous or Fragmented and Continuous are above 0.8, respectively. Time periods where none of these criterion are met are considered unclassified.

@@

\subsection*{Replay distance from animal}
Replay distance from the animal is defined as the shortest path distance along the track graph between the animal's 2D position projected on to the track graph (see Linearization) and the MAP estimate of the posterior probability of position, $\argmax_{x_k} p(x_{k} \mid O_{1:T})$, which is defined as the center of the position bin with the greatest posterior value. The shortest path is found using Dijkstra's algorithm \cite{Dijkstranotetwoproblems1959} as implemented in NetworkX \cite{HagbergExploringNetworkStructure2008}.

@@

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=1.0\linewidth]{figures/Figure1-supplemental1/Figure1_v3_supplemental1}
\caption{The model is robust to change of the probability of persisting in the same dynamic for a wide range of plausible expected durations (25 ms to 150 ms). \textbf{(A)} Each plot shows the probability of each dynamic on simulated data example from Figure 1 with a different diagonal value---which governs the probability of remaining in that state. The corresponding expected duration of staying in the state is listed as duration. The off-diagonal values---the probability of switching to one of the other dynamics---are set to be equally likely with the remainder of the probability, as in Figure \ref{1}D. The diagonal increases from left to right, top to bottom, until the case where the diagonal is one and the off-diagonal is zero---i.e. the case where there is no probability of switching to another state. Shaded regions correspond to the classification as in Figure \ref{1}. \textbf{(B)} The probability of position over time for each diagonal value. Conventions the same as in (A).}
\label{fig:Figure1-Figure supplement 1}
\end{figure*}

>>>>

\title{Modeling the movement dynamics of hippocampal replay reveals slower spatially coherent trajectories}
\shorttitle{Replay Dynamics}

@@

%TC:break Abstract
%the command above serves to have a word count for the abstract
\begin{abstract}
Representations of past and possible future experiences play a critical role in memory and decision-making processes. The hippocampus is known to express such representations during sharp-wave ripple (SWR) events, when ensembles of hippocampal place cells can be activated in time-compressed sequences consistent with prior experiences. Efforts to understand these "replay" events have typically imposed stringent criteria on SWR event significance, and thus analyze only a small subset of events (\texttildelow10-45\%). As a result, we lack a clear understanding of both the prevalence and the range of representational dynamics associated with replay. Here, we develop a state space model that uses a combination of movement dynamics of different speeds to capture the spatial content and time evolution of replay during SWR events. Application of this model reveals that over 60\% of SWR events contain slower spatially coherent trajectories or representations of single locations.

@@

We find that the majority of SWRs contain spatially coherent content; that is, the replay trajectories have no large jumps. Surprisingly, while the expected high speed replay events were identified, the most common category of events expressed representations that moved at much slower speeds, more consistent with stationary or slower moving trajectories, suggesting that high speed replay is not the most common manifestation of generative activity. These findings illustrate the power of state space models and provide a new understanding of the nature of hippocampal replay.

\section*{Results}
\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure1/Figure1_v6}
\caption{The model can capture different sequence dynamics on simulated data. \textbf{(A)} We construct a firing sequence of 19 simulated place cells that exhibits three different movement dynamics. For the first 60 ms, one cell fires repeatedly, representing one stationary location. For next 190 ms, the cells fire in sequence, representing a rapid continuous trajectory along the virtual track. Finally, for the last 30 ms, cells fire randomly, out of spatial order, representing an fragmented spatial sequence. \textbf{(B)} Like the standard decoder, the state space model uses estimates of cells' place fields from when the animal is moving and combines them with the observed spikes in (A) to compute the likelihood of position for each time step. \textbf{(C)} The prediction from the neural data is then combined with an explicit model of each movement dynamic, which determines how latent position can change based on the position in the previous time step. We show the probability of the next position bin for each movement dynamic model (color scale). Zero here represents the previous position. \textbf{(D)} The probability of remaining in a particular movement dynamic versus switching to another dynamic is modeled as having a high probability of remaining in a particular dynamic with a small probability of switching to one of the other dynamics at each time step. \textbf{(E)} The state space model uses the components in A-D over all time to decode the joint probability of latent position and dynamic. The joint probability can be summarized by marginalizing over latent position (left panel) to get the probability of each dynamic over time. The summed posterior across dynamics also provides an estimate of latent position over time (right panel). \textbf{(F)} The probability of each dynamic depends heavily on the speed of the trajectory, as we show using a range of simulated spiking sequences each moving at a constant speed. Each dot corresponds to the average probability of that dynamic for a given constant speed sequence. We use a 0.8 threshold (dotted line) to classify each sequence based on the dynamic or dynamics which contribute maximally to the posterior (shaded colors). 
}
\label{1}
\end{figure*}
\subsection*{Overview of the model}
We begin with an application of the model to simulated data, both to validate the model and to provide intuition (Figure \ref{1}). We simulated 19 Poisson spiking cells with Gaussian place fields on a 180 cm virtual linear track. Each place field has a 36 cm variance and a 15 Hz peak firing rate, which is spaced every 10 cm along the virtual track. We then construct an 280 ms spiking sequence and apply our decoding algorithm in Figure \ref{1}A. For the first 60 ms of this sequence, a single place cell fires repeatedly, resulting in the extended representation of a single location. For the second 190 ms of the sequence, the cells fire in sequential spatial order, representing a fast moving trajectory across the virtual linear track. For the last 30 ms of the sequence, the cells fire in an incoherent spatial order. These three firing patterns represent three different types of sequence dynamics that could be expressed during replay events, which we call stationary, continuous, and fragmented, respectively. The goal of our model is to characterize SWRs in terms of a mixture of these three dynamics at every time point.

@@

Next, we specify movement dynamic models for how latent position---the "mental" position of the animal represented by the cells---evolves over time for a given dynamic (Figure \ref{1}C). We do this by defining a state transition matrix, which defines how the latent position can move from the previous time step. Previous findings suggest that replay may exhibit at least three distinct types of movement dynamics: stationary \cite{JaiDistincthippocampalcorticalmemory2017, FarooqEmergencepreconfiguredplastic2019}, continuous \cite{DavidsonHippocampalReplayExtended2009}, and fragmented, which could correspond to both extended spatially incoherent representations and representations that jump from one position to another in a short time period \cite{PfeifferAutoassociativedynamicsgeneration2015}. We therefore define state transition models to capture each of these movement dynamics.

@@

Once we have specified the data and the movement dynamics models, we have fully specified the state space model. We use acausal decoding, meaning that we use all information from the past and future spikes, to estimate the joint posterior probability of position and dynamics (see Methods). With this, we can summarize the resulting posterior probability with two quantities: the probability of each movement dynamic over time (by integrating out position) and the probability of latent position over time, irrespective of movement dynamic (by summing over the movement dynamics; Figure \ref{1}E, left and right plot respectively).

An examination of the two summaries shown in Figure \ref{1}E reveals that that the model successfully captures the dynamics of the population spiking activity in Figure \ref{1}A. The stable firing of the one active neuron indicates a stationary representation, and accordingly, the probability of the stationary movement dynamic is high at the beginning of the simulated replay. A change in the data then drives a rapid transition to the continuous movement dynamic, reflecting the trajectory-like spatially sequential spiking from the population of cells. Subsequently, as the population activity becomes spatially incoherent, the fragmented movement dynamic dominates for the last third of the simulated event.

The specification of these movement dynamics allows the model to capture a wide range of movement speeds for the latent position. To demonstrate this, we applied the model to 10,000 simulated trajectories, each trajectory proceeding at a constant speed (Figure \ref{1}F) from 1 cm/s to 10,000 m/s. From this, we can see that not only are there regions of speed that correspond to our three movement dynamics being highly probable (where we define highly probable to be greater than or equal to 0.8 probability), there are intermediate speeds where two of the dynamics exhibit relatively high probability; and where the sum of two of the dynamics' probabilities exceeds 0.8. In this manuscript, we will refer to these as mixture dynamics. For example, when the stationary dynamic has a probability of 0.6 and the continuous has a probability of 0.4, we call this a stationary-continuous mixture (light blue, Figure \ref{1}G). By using this classification scheme, we can characterize the speed or set of speeds in a replay by its movement dynamics. To summarize, there are three movement dynamic models, each of which is assigned a probability over time. We classify these probabilities as being part of one of five speed categories: stationary, stationary-continuous-Mixtures, continuous, fragmented-continuous Mixtures, and fragmented.

@@

As mentioned above, the model is robust to the choice of probability of staying in a dynamic. To demonstrate this we decoded the spiking sequence in Figure \ref{1}A with different probabilities of staying in the same dynamic versus switching to another dynamic (Figure \ref{fig:Figure1-Figure supplement 1}). We found that for a large range of plausible probabilities of staying in one of the dynamics over a single 2 ms time step (between 0.96 and 0.993 which corresponds to an expected duration between 25 and 150 ms), the model still correctly identified the dynamic with high probability.

@@

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure3/Figure3_v5}
\caption{A-F. Six examples of SWRs with non-constant speed trajectories. Figure conventions are the same as in Figure \ref{2}. Filtered SWR ripple (150-250 Hz) trace from the tetrode with the maximum amplitude displayed above each example. \textbf{(A)} A SWR where the decoded position starts moving down the center arm away from the animal's position at the center well, slows down, and returns back. \textbf{(B)} A SWR where the decoded position persistently stays at the choice point (open circle) while the animal remains at the left well. \textbf{(C)} A SWR where the decoded position begins with stationary representation of the left well, then jumps to the middle of the right arm and proceeds up the right arm to the right well. \textbf{(D)} A SWR where the decoded position begins with stationary representation of the left well, jumps to the center arm, proceeds away from the center well, jumps to the right arm, proceeds back toward the center well, and then becomes fragmented. \textbf{(E)} A SWR where the decoded position begins in the left arm and persists at the choice point. \textbf{(F)} A SWR where the decoded position starts in the left arm toward the choice point, jumps to the right arm and proceeds back toward the choice point. \textbf{(G)} Each dot represents the percentage of SWRs for each day for that animal. A SWR is included in the numerator of the percentage if at any time it includes the classifications listed below the column. The denominator is listed in the x-axis label.
}
\label{3}
\end{figure*}

We next sought to validate the model on real hippocampal data. We fit the place fields of 31 cells recorded in hippocampus while the rat was performing a spatial alternation task on a W-shaped track, and then applied the decoding algorithm to a SWR with sequential population activity (Figure \ref{2}A, top panel). As expected, we observed that the probability of being in the continuous dynamic is high throughout this event, but the probability of being in a stationary dynamic was also noticeable at the beginning and end of the SWR (Figure \ref{2}A, middle panel). Using our speed classification scheme, this means that the speed of the replay starts slower---as a mixture of continuous and stationary dynamics---and then speeds up and slows down again. This is also evident in the posterior probability of linear position over time. This shows that the replay most likely travels down the center arm and up the right arm (Figure \ref{2}A, bottom panel). We can also see this when we project the maximum of the posterior of this trajectory (the most probable "mental" position) to 2D (Figure \ref{2}C) to better see the spatial trajectory on the maze. Importantly, when we apply the same model using the 2D position of the animal, we get a similar result as when we use the 1D linearized position (Figure \ref{fig:Figure2-Figure supplement 1}A).

@@

As expected, and providing further support for the validity of the method, we observed many replays that were classified as continuous throughout the entirety of the SWR as with the standard decoder(Figure \ref{fig:Figure2-Figure supplement 2}A-F). However, we also observed many events with spatially coherent content that did not have this structure. For example, we observed trajectories that started in one direction and reversed back to the original position (Figure \ref{3}A, Figure \ref{fig:Figure3-Figure supplement 1}B), trajectories that remained fixed in one position (Figure \ref{3}B, Figure \ref{fig:Figure3-Figure supplement 1}G), trajectories that jumped between arms and between dynamics (Figure \ref{3}C-F, Figure \ref{fig:Figure3-Figure supplement 1}F, H, I), and trajectories that were spatially incoherent throughout the SWR (Figure \ref{fig:Figure3-Figure supplement 1}A, D).

Using a 0.8 threshold, we were able to classify 89\% (23382 of 26160) of SWRs as containing at least one of the three dynamics or common dynamic mixtures. To ensure that this depended on the sequential firing of spatially tuned cells in the hippocampus, we resampled position with replacement for two recording sessions, shuffling the relationship between spiking and spatial information and then fitting the encoding model. We then decoded the same SWR events containing the same spikes. Only 9\% of the SWRs were classified in the shuffled datasets, a value that was significantly less than that seen for the real data (p=0.02 for recording session 1, p=0.02 for recording session 2, Figure \ref{fig:Figure3-Figure supplement 2}).

@@

To better compare our findings to previous work, we quantified the percentage of classified SWRs that contained continuous content, as would typically be analyzed when using the standard decoder. Here our findings were consistent with previous reports: in our datasets, 4511 of 23382 or 19\% of classified SWRs had time periods where the decoded position was classified as continuous (Figure \ref{3}G, 17\% of all SWRs). Thus, focusing on only high speed trajectories misses a large fraction of events where there is evidence for spatially coherent, but not continuous content.

@@

Our model is specified in the context of a latent position associated with different movement dynamics. This specification allows us to not only to classify events in terms of their dynamics, but also to quantify the model's certainty in each position estimate at each moment in time given the model parameters. To do so we can compute the cumulative spatial extent of the 95\% highest posterior density (HPD) region of the posterior linear position. Larger values of the cumulative HPD extent indicate the model is less certain about position, because the most probable decoded positions are distributed over more of the track at a given time point. In contrast, lower values indicate that the model is more certain about the estimate of position because the extent of the HPD is more concentrated and covers less of the track. Thus, the HPD provides a complementary measure of spatial coherence, and evaluating it allows us to verify that the events we defined as spatially coherent correspond to events where there is high certainty around the position estimates. Although this is certainly influenced by the dynamics imposed on the model via the movement state transition matrix, because the model is primarily driven by the data, it does not necessarily have to be the case that the two coincide. To show this, as before, we resampled position with replacement for two recording sessions, shuffling the relationship between spiking and spatial information and then fitting the encoding model. The shuffled decodes had much larger spatial coverage than the real data (XX cm vs XX cm, p=2.2e-16, one-sided Mann-Whitney-U, \ref{fig:Figure4-Figure supplement 1}).

@@

\subsection*{Many replay trajectories have slower movement speeds}
\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure5/Figure5_v4}
\caption{
Quantification of the SWRs by dynamic. \textbf{(A)} UpSet plot \cite{LexUpSetVisualizationIntersecting2014}---which is similar to a Venn diagram with more than 3 sets---of the most common sets of classifications within each SWR. Each row represents a classification and each column represents a set of classifications, where filled-in black dots with an edge between the dots indicates that multiple classifications are present in the SWR (at temporally distinct times). The sets are ordered by how often they occur as indicated by the bar plot above each category. The total percentage of each classification is indicated by the rotated bar plot on the left. \textbf{(B)} Box plot of the duration of each dynamic category within a SWR. The box shows the interquartile range (25-75\%) of the data and the whiskers are 1.5 times the interquartile range. Points outside of this are labeled as diamonds. \textbf{(C)} Box plot of the average distance between the latent position to the animal's position for each classification within the SWR. \textbf{(D)} Box plot of the average speed of the classification within the SWR (excluding classifications < 20 ms).\textbf{(E)} Kernel density estimate of the position of stationary trajectories on the W-track at least 30 cm away from the animal's position. The shaded region represents the density estimate while the black ticks represent the observed non-local stationary positions. \textbf{(F)} Box plot of the average tetrode multiunit spike rates for each dynamic category within each SWR (excluding classifications < 20 ms).
}
\label{5}
\end{figure*}

@@

Previous work reported that less than half of SWRs or high-population-activity events contained replay \cite{KarlssonAwakereplayremote2009, FosterReversereplaybehavioural2006, DavidsonHippocampalReplayExtended2009}. This could lead one to the assumption that most events are not spatially coherent or meaningful. Our results indicate that this is not the case: in our dataset, the large majority of SWRs (85\%) contained spatially coherent content, as defined by some combination of stationary or continuous dynamics. These events included decoded representations that were highly spatially concentrated. This was in stark contrast to unclassified events or events that included fragmented dynamics, which were associated with highly distributed decoded spatial representations.

@@

Second, we used clusterless decoding, which enabled us to take advantage of information from neurons farther away from the tetrodes and not just those that could be cleanly and somewhat subjectively separated \cite{ChenBayesiannonparametricmethods2016, KloostermanBayesiandecodingusing2014, DengRapidclassificationhippocampal2016}. We note that more fully automated modern spike sorting algorithms (such as MountainSort \cite{ChungFullyAutomatedApproach2017}) in combination wiht better recording technology may eliminate some of this gap between sorted spikes and clusterless decoding, by removing experimenter subjectivity involved in spike sorting and by more accurately identifying neurons.

Third, we built an explicit, but flexible model that allowed us to identify not just one dynamic, but multiple dynamics consistent with different speeds of motion. And we use these dynamics to describe the SWR events, rather than declaring the events as significant or not. Fourth, our model avoids assumptions about temporal bins and does not make assumptions about the linear progression of events. Instead, the model can use any size of bin and allows for movement in any direction. Finally, we used an acausal estimator of latent position, which allowed us to identify the specific timing of transitions in dynamics at a fine timescale. Our method allows us to identify periods of slower moving events that

@@

\subsection*{Probability of the dynamics}
The probability of the dynamics is a quantity that indicates how much that dynamic is contributing to the posterior at a given time. We estimate the probability of the dynamics by integrating out position from the joint posterior:
$$Pr(I_{k} \mid O_{1:T}) = \int p(x_{k}, I_{k} \mid O_{1:T}) dx_{k}$$
As in other calculations, the integral is approximated with a Riemann sum.

@@

\subsection*{Classification of Dynamics}
We used a threshold of 0.8 to classify the probability of each state $Pr(I_{k} \mid O_{1:T})$ into 5 categories. Time periods during sharp wave ripples are labeled as stationary, continuous, or fragmented when the probability of each individual state is above 0.8. Time periods are labeled as stationary-continuous-Mix or fragmented-continuous-Mix when the sum of stationary and continuous or fragmented and continuous are above 0.8, respectively. Time periods where none of these criterion are met are considered unclassified.

@@

\subsection*{Distance from animal}
The distance of the decoded position from the animal's is defined as the shortest path distance along the track graph between the animal's 2D position projected on to the track graph (see Linearization) and the MAP estimate of the posterior probability of position, $\argmax_{x_k} p(x_{k} \mid O_{1:T})$, which is defined as the center of the position bin with the greatest posterior value. The shortest path is found using Dijkstra's algorithm \cite{Dijkstranotetwoproblems1959} as implemented in NetworkX \cite{HagbergExploringNetworkStructure2008}.

\subsection*{Estimation of Speed}
In order to estimate the speed over time, we first found the MAP estimate of the posterior probability of position, $\argmax_{x_k} p(x_{k} \mid O_{1:T})$, which is defined as the center of the position bin with the greatest posterior value. We then computed the first derivative using the Python library Numpy's gradient function, which computes the difference in the forward and backward time direction and then averages them for all points except the boundaries. We then smoothed this quantity with a small Gaussian (2.5 ms standard deviation) and then averaged this for the classified time points.

@@

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=1.0\linewidth]{figures/Figure1-supplemental1/Figure1_v3_supplemental1}
\caption{The model is robust to change of the probability of persisting in the same dynamic for a wide range of plausible expected durations (25 ms to 150 ms). \textbf{(A)} Each plot shows the probability of each dynamic on simulated data example from Figure 1 with a different diagonal value---which governs the probability of remaining in that dynamic. The corresponding expected duration of staying in the dynamic is listed as duration. The off-diagonal values---the probability of switching to one of the other dynamics---are set to be equally likely with the remainder of the probability, as in Figure \ref{1}D. The diagonal increases from left to right, top to bottom, until the case where the diagonal is one and the off-diagonal is zero---i.e. the case where there is no probability of switching to another dynamic. Shaded regions correspond to the classification as in Figure \ref{1}. \textbf{(B)} The probability of position over time for each diagonal value. Conventions the same as in (A).}
\label{fig:Figure1-Figure supplement 1}
\end{figure*}
